---
title: "hERG"
author: "Jake Smith"
date: "2/4/2019"
output: html_document
  latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(18438)

require(ggplot2)
require(dplyr)
require(tidyr)
```

The hERG activity data was collected from the PubChem database, PubChem AID 588834, "QHTS Assay For Small Molecule Inhibitors Of The Human HERG Channel Activity". Molecular descriptors were calculated using KNIME-CDK version 1.5.600.v201711101057.

```{r}
# import the hERG dataset
hERG <- read.csv("hERG_descriptors.csv")

# remove Score and activators
hERG <-  filter(hERG, !Phenotype == "Activator") %>% select(-Score, -row.ID)

# convert potency to pIC50, replace NA with 0
hERG <- mutate(hERG, Potency = -1 * log(Potency * 1e-6, base = 10))
hERG <- mutate(hERG, Potency = case_when(is.na(Potency) ~ 0, !is.na(Potency) ~ Potency))

# re-encode phenotype as success/fail
hERG <- mutate(hERG, Phenotype = case_when(Phenotype == "Inactive" ~ 0, Phenotype == "Inhibitor" ~ 1))
```

```{r}
## plot observed hERG activity as a function of the chemical space

# generate principle components from the major numeric descriptors
# log transform predictors where necessary to get closer to normal
hERG.PCA <- prcomp(formula = ~ log(Bond.Polarizabilities) + log(VABC.Volume.Descriptor) + log(Topological.Polar.Surface.Area + 1) + log(Molecular.Weight) + XLogP + log(Eccentric.Connectivity.Index), data = hERG, center = TRUE, scale. = TRUE)

# plot the chemical space using the principle components, color by activity score
chemspace <- ggplot(data = hERG) + geom_point(aes(x = hERG.PCA$x[,1], y = hERG.PCA$x[,2], color = Phenotype))

# show plot
chemspace
```

```{r}
# plot histograms for each of the explanatory variables in the hERG dataset
ggplot(data = (select(hERG, -Potency) %>% gather(key, value))) + facet_wrap(~ key, scale = "free") + geom_histogram(aes(value), bins = 50)
```
```{r}
# fit a logistic regression for phenotype as a function of all other variables
glm.v1 <- glm(Phenotype ~ ., data = select(hERG, -Potency), family = binomial)

# output a summary
summary(glm.v1)
```

```{r}
# calculate VIF for the explanatory variables
car::vif(glm.v1)
```

Initially remove hydrogen bond donors, rotatable bond count, and formal charge -> low significance descriptors with minimal covariance to confuse the evaluation

Remove molecular weight -> VABC volume is a more detailed measure of the same general property.

XLogP gets to remain in the model, as it is a commonly used heuristic in medicinal chemistry whose exclusion would be notable.

```{r}
# remove variables and refit
glm.v2 <- glm(Phenotype ~ Aromatic.Atoms.Count + Bond.Polarizabilities + VABC.Volume.Descriptor + Hydrogen.Bond.Acceptors + Topological.Polar.Surface.Area + XLogP + SP3.Character + Eccentric.Connectivity.Index + Amine, data = select(hERG, -Potency), family = binomial)

# output a summary
summary(glm.v2)
```

```{r}
# calculate VIF for the explanatory variables
car::vif(glm.v2)
```

Remove bond polarizabilities and hydrogen bond acceptors -> high VIF, conceptually related to several other remaining descriptors

```{r}
# remove variables and refit
glm.v3 <- glm(Phenotype ~ Aromatic.Atoms.Count + VABC.Volume.Descriptor + Topological.Polar.Surface.Area + XLogP + SP3.Character + Eccentric.Connectivity.Index + Amine, data = select(hERG, -Potency), family = binomial)

# output a summary
summary(glm.v3)
```

```{r}
# calculate VIF for the explanatory variables
car::vif(glm.v3)
```

Finally, we will try to remove eccentric connectivity index. While it does appear to be contributing significantly, it is quite damaging to the interpretability of the model.

Similarly, while both are contributing significantly, aromatic atoms count and sp3 character are closely related molecular properties. In the interest of improving interpretability, we will remove sp3 character, the less accessible of the two.

```{r}
# remove variables and refit
glm.v4 <- glm(Phenotype ~ Aromatic.Atoms.Count + VABC.Volume.Descriptor + Topological.Polar.Surface.Area + XLogP + Amine, data = select(hERG, -Potency), family = binomial)

# output a summary
summary(glm.v4)
```

```{r}
# calculate BIC for the fitted models
BIC(glm.v1, glm.v2, glm.v3, glm.v4)
```

```{r}
## cross-validation for the intermediate models

# ten-fold partition the hERG dataset
logit.folds <- caret::createFolds(hERG$Phenotype, k = 10)

# add new columns to hERG to hold predictions
hERG$Prediction1 <- NA
hERG$Prediction2 <- NA
hERG$Prediction3 <- NA
hERG$Prediction4 <- NA

# define a restricted dataset to run regressions on
hERG.restict <- select(hERG, -Potency, -Prediction1, -Prediction2, -Prediction3, -Prediction4)

# perform ten-fold cross-validation on the three logistic regressions
for (i in 1:10) {
  # fit the models on the 9/10 sets
  glm1 <- glm(Phenotype ~ ., data = hERG.restict[-logit.folds[[i]], ], family = binomial)
  
  glm2 <- glm(Phenotype ~ Aromatic.Atoms.Count + Bond.Polarizabilities + VABC.Volume.Descriptor + Hydrogen.Bond.Acceptors + Topological.Polar.Surface.Area + XLogP + SP3.Character + Eccentric.Connectivity.Index + Amine, data = hERG.restict[-logit.folds[[i]], ], family = binomial)

  glm3 <- glm(Phenotype ~ Aromatic.Atoms.Count + VABC.Volume.Descriptor + Topological.Polar.Surface.Area + XLogP + SP3.Character + Eccentric.Connectivity.Index + Amine, data = hERG.restict[-logit.folds[[i]], ], family = binomial)
  
  glm4 <- glm(Phenotype ~ Aromatic.Atoms.Count + VABC.Volume.Descriptor + Topological.Polar.Surface.Area + XLogP + SP3.Character + Amine, data = hERG.restict[-logit.folds[[i]], ], family = binomial)

  # predict with the 1/10 sets
  glm1.pred <- predict(glm1, select(hERG, -Potency)[logit.folds[[i]], ])
  glm2.pred <- predict(glm2, select(hERG, -Potency)[logit.folds[[i]], ])
  glm3.pred <- predict(glm3, select(hERG, -Potency)[logit.folds[[i]], ])
  glm4.pred <- predict(glm4, select(hERG, -Potency)[logit.folds[[i]], ])

  # store predicted values back in hERG
  hERG[logit.folds[[i]], ] <- cbind(hERG[logit.folds[[i]], 1:15], glm1.pred, glm2.pred, glm3.pred, glm4.pred)
}

# calculate AUC
pROC::auc(response = hERG$Phenotype, predictor = hERG$Prediction1, threshold = 0.5)
pROC::auc(response = hERG$Phenotype, predictor = hERG$Prediction2, threshold = 0.5)
pROC::auc(response = hERG$Phenotype, predictor = hERG$Prediction3, threshold = 0.5)
pROC::auc(response = hERG$Phenotype, predictor = hERG$Prediction4, threshold = 0.5)
```

Trimming down the explanatory variables included in the model has cost us in terms of BIC, but cross-validation shows minimal loss in predictive ability.

```{r}
## plot terms versus logit(Phenotype)
# pull out the terms included in the glm.v4
glm.v4.terms <- names(glm.v4$coefficients)[2:length(glm.v4$coefficients)]

# plot each explanatory variable versus logit(Pr)
ggplot(data = hERG %>% select(Prediction4, glm.v4.terms) %>% gather("key", "value", -Prediction4), 
       aes(x = value, y = Prediction4)) + geom_point() + geom_smooth(method = "loess", se = FALSE) + 
  facet_wrap(~key, scales = "free")
```

Aromatic atoms count looks like a fairly linear relationship. TPSA and VABC each look significantly curved, and XLogP looks cubic. We will try introducing higher order terms for those variables.

```{r}
## cross-validation for models with higher order terms

# ten-fold partition the hERG dataset
logit.folds <- caret::createFolds(hERG$Phenotype, k = 10)

# add new columns to hERG to hold predictions
hERG$Prediction5 <- NA

# define a restricted dataset to run regressions on
hERG.restict <- select(hERG, -Potency, -Prediction1, -Prediction2, -Prediction3, -Prediction4, -Prediction5)

# perform ten-fold cross-validation on the three logistic regressions
for (i in 1:10) {
  # fit the models on the 9/10 sets
  glm5 <- glm(Phenotype ~ Aromatic.Atoms.Count + VABC.Volume.Descriptor + I(VABC.Volume.Descriptor^2) + Topological.Polar.Surface.Area + I(Topological.Polar.Surface.Area^2) + XLogP + I(XLogP^3) + Amine, data = hERG.restict[-logit.folds[[i]], ], family = binomial)

  # predict with the 1/10 sets
  glm5.pred <- predict(glm5, select(hERG, -Potency)[logit.folds[[i]], ])

  # store predicted values back in hERG
  hERG[logit.folds[[i]], ] <- cbind(hERG[logit.folds[[i]], 1:19], glm5.pred)
}

# calculate AUC
pROC::auc(response = hERG$Phenotype, predictor = hERG$Prediction5, threshold = 0.5)
```

```{r}
# fit the model with higher order terms added
glm.v5 <- glm(Phenotype ~ Aromatic.Atoms.Count + VABC.Volume.Descriptor + I(VABC.Volume.Descriptor^2) + Topological.Polar.Surface.Area + I(Topological.Polar.Surface.Area^2) + XLogP + I(XLogP^3) + Amine, data = select(hERG, -Potency), family = binomial)

# output a summary
summary(glm.v5)
```

